---
title: "FinalExam174_2025"
author: "Stephen R. Proulx"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(rethinking)
library(bayesplot)
#source("../helper.R")

``` 





## Short Answer

### Question 1 (5 pts)

Consider this dataset where $y$ is the number of eggs laid by individual fruit flies who are put into vials with $x$ microliters of yeast paste. 
```{r}
data <- tibble( y = c(35,43,54,33,65,54,72,55,66), x = c(5,5,5,10,10,10,15,15,15))
```

What's wrong with this model specified here? Re-write the `ulam` model and explain what you changed and why
```{r}
model <- ulam(alist(
  y ~ dbinom(total, mu),
  mu <- a + b * x,
  a ~ dnorm(0,3),
  b ~ dnorm(0,1)
) ,  data=data)
```




### Question 2 (10 pts)
Consider a dataset where researchers measured body length (cm) in lizards. The data are coded:

| sex    | color | S | C |
|--------|-------|----|----|
| Male   | Brown | 0  | 0  |
| Male   | Green | 0  | 1  |
| Female | Brown | 1  | 0  |
| Female | Green | 1  | 1  |

So S=0 for male and S=1 for female lizards, and C=0 for brown and C=1 for green lizards.

Now imagine adding an additional index variable I to identify the sex/color combination for each lizard, using the mapping:

I=1 for (S=0, C=0)
I=2 for (S=0, C=1)
I=3 for (S=1, C=0)
I=4 for (S=1, C=1)

Consider these models for length

Model 1
$$
 y \sim \mathrm{normal}(\mu,\sigma) \\
 \mu = a + b_S S + b_C C
 
$$
Model 2
$$
 y \sim \mathrm{normal}(\mu,\sigma) \\
 \mu = a + b_S S + b_C C + b_{SC} S C
 
$$
Model 3
$$
 y \sim \mathrm{normal}(\mu,\sigma) \\
 \mu = a[I]
$$
#### (A) Which of these models can be considered models with interaction? Explain why.


#### (B) Consider these values for the means of categories below: 


sex   color    mean length    
----- -------  --------------
M      B         13
M      G         15
F      B         16
F      G         20

For each model, either choose parameter values that reproduce these four category means exactly, or explain why that model cannot do so. Briefly explain what you observe.



### Question 3 (10 pts)
Consider this dataset:

```{r}
data <- tibble( x=seq(1,10),y=c(4.4,6.0,7,8,8,7,5,3,2,5))

ggplot(data,aes(x=x,y=y))+ 
  geom_point()+
  ylim(0,10)
```




Here are two potential models to fit these data (you do not need to run these to answer the questions):
```{r}
m.1 <- ulam(alist(
  y ~ dnorm(mu, sigma),
  mu <- a + b1*x+ b2*x^2,
  a ~ dnorm(5,2.5),
  b1 ~ dnorm(0,1),
  b2 ~ dnorm(0,1),
  sigma ~ dexp(1)),
  data=data, log_lik = TRUE
)
```

```{r}
m.2 <- ulam(alist(
  y ~ dnorm(mu, sigma),
  mu <- a + b1*x + b2*x^2 + b3*x^3,
  a ~ dnorm(5,2.5),
  b1 ~ dnorm(0,1),
  b2 ~ dnorm(0,1),
  b3 ~ dnorm(0,1),
  sigma ~ dexp(1)),
  data=data, log_lik = TRUE
)
```

#### (A) 
After fitting the models you can compute the probability of observing the data given the parameters. Which model do you expect to fit the observed data best, and why? How would you define "fit better"?



#### (B) What if new data is added?

Now consider this additional point is discovered and added to the dataset.

Would this change which model you would prefer?

Explain your reasoning in terms of predicting this new data point rather than re-fitting the model with the new datum included.

What method from this course could you use to formally compare the predictive performance of these models?

```{r}

new_data <-  tibble( x=c(10), y= c(0.0) )

ggplot(data,aes(x=x,y=y))+ 
  geom_point()+
  geom_point(data=new_data,color="red")+
  ylim(0,10)
```


## Analysis Challenge


The paper on bacterial coevolution discusses prey bacteria that have been experimentally evolved in the presence of a predator. These evolved bacteria were then assayed for their fitness, and this was done in 4 environments. In the "no predator" environment the bacteria are allowed to grow without a predator. In the "coevolved predator" treatment they are being consumed by the predator strain that was present at the end of the coevolution experiment. In the "ancestral predator" treatment they are being consumed by the ancestral predator that has not undergone coevolution. In the "control predator" treatment a predator that was evolved in the absence of prey is included. 

We would like to ask whether or not the evolved prey bacteria respond differently to the 4 treatments. But first you will set up some simpler models that explore the overall variability among measurements without using the treatment variables.


In the experimental design, there were 12 separate evolved populations, and each population was assayed 3 times. The 12 evolved populations come from 3 different subclones (4 populations from each) and are either antibiotic resistant (resistance=="r") or not.


First we load the data:
```{r}
data<-read.csv("PreyCompetitionResults.csv")%>%
  as_tibble()%>%
  mutate(treatment_code=(treatment=="coevolved predator")*1 +(treatment=="control predator")*2+(treatment=="ancestral predator")*3+(treatment=="no predator")*4,
         resistant_code=(resistance=="r")*1+(resistance=="s")*2,log_rel_fitness=log(rel_fitness))%>%
  mutate(obs_id =(population*100+treatment_code))%>%
  mutate(observation=as.numeric(as.factor(obs_id)))  

```

Note that I created a column called "observation" which uniquely identifies each population by treatment. i.e. population 1 with the coevolved predator is coded as observation group 1, and has three replicate observations. So each population–treatment combination will be treated as a single observation group with three replicate measurements.


And we can start by visualizing the data, showing the growth rates for all 12 populations. You can see considerable variability in the growth rates among assays and between populations. 
```{r}
ggplot(data,aes(x=population,y=log_rel_fitness,color=as.factor(subclone))) +
  geom_point()+
  facet_wrap(~treatment)+
  scale_x_continuous(breaks = 1:12)


```



### Q4  create a complete pooling model for all the observations with no predictive factors. (10 pts)
The dataset has 143 observations, this comes from 12 populations assayed in 4 environments and replicated 3 times ($12*4*3=144$, with one observation "missing"). Here is a way to specify the model:
$$
\mathrm{log\ fitness} \sim \mathrm{Normal}(\mu,\sigma) \\
\mu \sim \mathrm{Normal}(0,2) \\
\sigma \sim \mathrm{Exponential}(1)
$$

(A) Explain why this specification for the likelihood and for the priors is appropriate. What makes this a complete pooling model?




(B) Specify the model using `ulam` and report the output. Explain how you can assess if the MCMC algorithm adequately sampled the posterior distribution.






### Q5 make no-pooling model at the level of evolved populations (10 pts)
To do this we have to group the replicate assays of each evolved population in each predator environment. I have already created the column "observation" for this. 

(A) Write out the no-pooling model where we assign a mean log growth rate for each observation group.



(B) Specify the model using `ulam` and report the output, including all parameters and hyperparameters. How do the Pooling and No Pooling models compare in their assessment of variability at the level of the relative fitness observation?




### Q6 make partial pooling model at the level of evolved populations (10 pts)

(A) Write out the partial-pooling model where we assign a mean log growth rate for each observation group, but that observation group value has a prior that includes a standard deviation parameter. 



(B) Specify the model using `ulam` and report the output. Is this any different in terms of sigma than the No Pooling model? 





### Q7 make partial pooling model with predator treatment as factor. (20 pts)


(A) Write out a model where the effect of observation group (population–treatment combination) is a partial-pooling model (i.e. a random effect) and where there is an effect of predator treatment.



(B) Specify the model using `ulam` and report the precis output. 



(C) Compute contrasts for the treatment effects. Because there are 4 treatment types, there are 6 contrasts (1-2, 1-3, 1-4, 2-3, 2-4, 3-4).  How can you interpret these contrasts to tell us how the bacteria evolved under the four different experimental treatments?



(D) Compare the four models using `compare`. Explain what you conclude based on the WAIC scores. Interpret the pWAIC values. 





### Q8 Add in additional co-factors (5 pts)
The populations are also defined by which subclone they were derived from and whether or not they are antibiotic resistant.

(A) Write out a model that extends the model from Q7 and  includes subclone identity and resistance.



(B) Use `ulam` to run this model and report the `precis`



(C) Compute the contrasts for treatment (the same six contrasts as in Q7) using the posterior samples for this model











