---
title: "Multilevel III"
author: "Stephen R. Proulx"
output: pdf_document
---

```{r setup_hier_normal_ulam, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(rethinking)
source("../helper.R")

```



```{r}
cmdstanr::set_cmdstan_path("~/.cmdstan/cmdstan-2.34.1")
```



# Today's objectives:  
* describe hierarchical models with normal likelihoods
* See how we can write equivalent models with the same likelihood (but different priors)
* Include more than one heirarchical level




# Data simulation for normal data in replicates

Here we simulate data where there is a treatment and replicates. The treatment effect is quite large, but we may have data where the "treatment" is hidden from us. i.e. it is some unknown variability in our system.
```{r}
mu_low=10;
mu_high=20;
sigma_noise=1;
sigma_obs=3;
set.seed(42)


group_effect=tibble(replicate=seq(1:10),delta_rep = rnorm(10,mean=0,sd=sigma_noise))

mystery_data <- tibble(index=seq(1:100),replicate=rep(1:10,10),treat=rep(rep(1:2,10),5))%>%
  mutate(mu_treat = (treat==1)*mu_low+(treat==2)*mu_high,
         delta_obs=rnorm(n(),mean=0,sd=sigma_obs)) %>%
  left_join(group_effect)%>%
  mutate(y=mu_treat+delta_obs+delta_rep)

mystery_data_complete<- mystery_data

mystery_data<- select(mystery_data,y,treat,replicate)
```


```{r}
ggplot(mystery_data,aes(x=as.factor(replicate),y=y,group=as.factor(replicate)))+geom_boxplot()
```

## Hierarchical model, variable intercept (centered)

Assume for now that we haven't measured the treatment and do not have it in our dataframe. We can write a heirarchical variable intercept model :

$$
y_i \sim \text{Normal}(\mu_i,\sigma_{\text{obs}})\\
\mu_i = \mu_{\text{replicate}[i]}\\
\mu_r \sim \text{Normal}(\bar{\mu}, \sigma_{\text{rep}}) \\
\bar{\mu} \sim \text{Normal}(10,10)\\
\sigma_{\text{rep}} \sim \text{Exponential}(1)
$$

We can fit this model and observe the parameters from the posterior, it does a good job of capturing the mean, and suggests replicates are pretty variable. 
```{r}
m_normal_H1 <- ulam(
  alist(
    y ~ dnorm(mu,sigma),
    mu <-mu_rep[replicate],
    mu_rep[replicate] ~ dnorm(mubar,sigma_rep),
    mubar ~ dnorm(10,10),
    sigma_rep ~ dexp(1),
    sigma ~ dexp(1)
    ),
  data=mystery_data, chains=4, cores=10, iter=10000, log_lik = TRUE
  )
```

```{r}
precis(m_normal_H1,depth=2)
```

## Hierarchical model, variable intercept (non-centered)

Another way of writing it is

$$
y_i \sim \text{Normal}(\mu_i,\sigma_{\text{obs}}) \\ 
\mu_i = \bar{\mu} + \Delta_{\mu}_{\text{replicate}[i]}\\
 \Delta_{\mu}_{\text{r} \sim \text{Normal}(0, \sigma_{\text{rep}}) \\
\bar{\mu} \sim \text{Normal}(10,10)\\
\sigma_{\text{rep}} \sim \text{Exponential}(1)
$$


We can fit this model and observe the parameters from the posterior, it does a good job of capturing the mean, and suggests replicates are pretty variable. 
```{r}
m_normal_H2 <- ulam(
  alist(
    y ~ dnorm(mu,sigma),
    mu <-mubar+Delta_mu[replicate],
    Delta_mu[replicate] ~ dnorm(0,sigma_rep),
    mubar ~ dnorm(10,10),
    sigma_rep ~ dexp(1),
    sigma ~ dexp(1)
    ),
  data=mystery_data, chains=4, cores=10, iter=10000, log_lik = TRUE
  )
```

```{r}
precis(m_normal_H2,depth=2)
```
We can go one more step to abstract further and allow us to sample from a unit normal:
```{r}
m_normal_H3 <- ulam(
  alist(
    y ~ dnorm(mu,sigma ),
    mu <-mubar+Delta_mu[replicate]*sigma_rep,
    Delta_mu[replicate] ~ dnorm(0,1),
    mubar ~ dnorm(10,10),
    sigma_rep ~ dexp(1),
    sigma  ~ dexp(1)
    ),
  data=mystery_data, chains=4, cores=10, iter=10000, log_lik = TRUE
  )
```

```{r}
precis(m_normal_H3,depth=2)
```
# Heirarchichal model with "random" treatment effects

This is expressed as a multi-level hierarchical model.


$$
y_i &\sim \text{Normal}(\mu_{r[i]}, \sigma_{\text{obs}}) \\
\mu_r &\sim \text{Normal}(\mu_{t[r]}, \sigma_{\text{rep}})  \\
\mu_t &\sim \text{Normal}(\bar{\mu}, \sigma_{\text{treat}})  \\
\bar{\mu} &\sim \text{Normal}(10,10) \\
\sigma_{\text{obs}} &\sim \text{Exponential}(1) \\
\sigma_{\text{rep}} &\sim \text{Exponential}(1) \\
\sigma_{\text{treat}} &\sim \text{Exponential}(1)
$$



```{r}
treat_lookup <- mystery_data %>%
  distinct(replicate, treat) %>%
  arrange(replicate) %>%
  rename(treat_of_rep = treat)
stan_data_nested <- list(
  y = mystery_data$y,
  replicate = as.integer(mystery_data$replicate),
  treat_of_rep = as.integer(treat_lookup$treat_of_rep),
  R = as.integer(n_distinct(mystery_data$replicate)),
  T = as.integer(n_distinct(mystery_data$treat))
)
```


```{r}

m_normal_nested <- ulam(
  alist(
    y ~ dnorm(mu_rep[replicate], sigma_obs),
    vector[R]:mu_rep ~ dnorm(mu_treat[treat_of_rep], sigma_rep),
    vector[T]:mu_treat ~ dnorm(mubar, sigma_treat),
    mubar ~ dnorm(10, 10),
    sigma_obs ~ dexp(1),
    sigma_rep ~ dexp(1),
    sigma_treat ~ dexp(1)
  ),
  data = stan_data_nested,
  chains = 4,
  cores = 4,
  iter = 2000,
  log_lik = TRUE
)
 
```

```{r}
precis(m_normal_nested,depth=2)
```



```{r}
m_normal_nested_treat <- ulam(
  alist(
    y ~ dnorm(mu, sigma),
    mu <- mu_treat[treat] + delta_rep[replicate] ,
    delta_rep[replicate]  ~ dnorm(0, sigma_rep),
    mu_treat[treat] ~ dnorm(mubar, sigma_treat),
    mubar ~ dnorm(10,10),
    sigma ~ dexp(1),
    sigma_rep ~ dexp(1),
    sigma_treat ~ dexp(1)
  ),
  data = mystery_data,
  chains = 4,
  cores = 4,
  iter = 2000,
  log_lik = TRUE
)
```

```{r}
precis(m_normal_nested_treat,depth=2)
```




# data in nested clusters/replicates

```{r}
set.seed(42)

# parameters
n_cluster <- 15
n_rep <- 3
n_obs_per_rep <- 100

mu_bar <- 10
sigma_cluster <- 5
sigma_rep <- 1
sigma_obs <- 3

# cluster-level effects
cluster_effect <- tibble(
  cluster = 1:n_cluster,
  delta_cluster = rnorm(n_cluster, 0, sigma_cluster)
)

# replicate-level effects nested within clusters
rep_effect <- tibble(
  cluster = rep(1:n_cluster, each = n_rep),
  replicate = 1:(n_cluster * n_rep),
  delta_rep = rnorm(n_cluster * n_rep, 0, sigma_rep)
)

# observation-level data
sim_data <- tibble(
  index = 1:(n_cluster * n_rep * n_obs_per_rep),
  replicate = rep(rep_effect$replicate, each = n_obs_per_rep)
) %>%
  left_join(rep_effect, by = "replicate") %>%
  left_join(cluster_effect, by = "cluster") %>%
  mutate(
    mu = mu_bar + delta_cluster + delta_rep,
    y = rnorm(n(), mu, sigma_obs)
  )

myster_data_cluster <- select(sim_data,y,replicate,cluster)

```


```{r}
ggplot(sim_data, aes(x = factor(replicate), y = y)) +
  geom_boxplot() +
  facet_wrap(~cluster, scales = "free_x") +
  labs(
    x = "Replicate",
    y = "Observed value (y)",
    title = "Replicates nested within clusters"
  ) +
  theme_bw()
```


```{r}
m_normal_clustered <- ulam(
  alist(
    y ~ dnorm(mu, sigma),
    mu<- mubar + delta_cluster[cluster]+delta_rep[replicate],
    delta_cluster[cluster] ~ dnorm(0,sigma_cluster),
    delta_rep[replicate]  ~ dnorm(0, sigma_rep),
    mubar ~ dnorm(10, 10),
    sigma ~ dexp(1),
    sigma_rep ~ dexp(1),
    sigma_cluster ~ dexp(1)
  ),
  data = myster_data_cluster,
  chains = 4,
  cores = 4,
  iter = 2000,
  log_lik = TRUE
)
```


```{r}
precis(m_normal_clustered,depth=2)
```




